\section{Related Work} \label{integr-related-work-sect}
Flexible methods that capture important source-specific features, but simultaneously model shared features between data sources have been recently proposed. 

\citet{Ray2012} proposed a Bayesian latent variable model for integrating multiple heterogeneous data, where the latent space was factorized into a shared component and a data source-specific component, employing a Beta-Bernoulli process \citep{Griffiths2005}. The proposed approach was demonstrated for analysis of multiple types of genomic data for ovarian cancer patients, collected from the Cancer Genome Atlas (TCGA) project.

\citet{Rey2012} implemented a dependency-seeking clustering approach, based on \emph{copula} mixture model, which considers the associations across the data sources when performing clustering. \citet{Savage2010} performed a mixture modelling approach, and used a Hierarchical Dirichlet Process \citep{Ferguson1973, Escobar1995} to simultaneously cluster transcription factor binding and gene expression data. \citet{Kirk2012} proposed a more general framework for performing integrative modelling for two or more data sources, using a Dirichlet-multinomial allocation mixture model \citep{Green2001}. They refer to their method as Multiple Dataset Integration (MDI), and it performs separate clustering while it simultaneously models the pairwise dependence between clusterings. 

Finally, \citet{Lock2013} developed Bayesian Consensus Clustering (BCC), which is an extension of simple mixture models to account for multiple data sources. BCC simultaneously models the dependence and the heterogeneity of the data sources, by assuming that there is a separate clustering of the objects for each data source, but these source-specific clusterings adhere loosely to an overall consensus clustering. \citet{Lock2013} showed that MDI method is equivalent to BCC (by performing parameter substitution) when the number of clusters and the number of data sources is equal to 2, since the MDI models pairwise dependence between clusterings, whereas BCC models the adherence to an overall consensus clustering and thus is more general.